\section{Experiments}
\label{sec:experiments}

This section evaluates PhyST-Net on representative industrial wind-power forecasting tasks where both predictive accuracy and operational feasibility are important. To validate the model's capacity to handle physical constraints, geographic relations, and external weather-driven forcing, we instantiate the framework on distributed wind-power generation networks as a concrete benchmark family. Following the formulation in Section~\ref{sec:problem_definition}, each sample contains historical turbine-level target states, available historical covariates, forecast-horizon exogenous drivers, and graph metadata. The target is active power over the future horizon. Unless otherwise stated, all reported errors are computed on the held-out test split after inverse transformation to physical units.

\subsection{Experimental Protocol}
\label{sec:experimental_protocol}

We evaluate two operational horizons: short-term and medium-term forecasting. These settings are selected because the physical spatio-temporal interface is expected to be most informative when exogenous forcing, graph propagation, and constraint-aware decoding jointly affect the forecast. The main comparison therefore focuses on short- and medium-horizon tasks on three wind-power datasets with different spatial scales: SDWPF, HoT, and Kelmarsh.

SDWPF is a Chinese wind-farm dataset from Shandong with 134 turbines sampled every 15 minutes over more than two years. The loader supports both the historical HDF5 files and the normalized CSV layout, including turbine-level power targets, SCADA measurements, numerical weather prediction (NWP) variables, and turbine positions. The current SDWPF configuration uses two primary meteorological features, such as average wind speed and average wind direction, together with standardized coordinates for spatial encoding. HoT is the Hill of Towie wind-farm dataset from Scotland, containing 21 turbines, 192794 timestamps, five exogenous variables, and 10-minute records from 2020-01-01 to 2024-09-01; the target missing rate is 1.83\%. Kelmarsh is a UK wind-farm dataset with six turbines, 157536 timestamps, 12 exogenous variables, and 10-minute records from 2016-01-03 to 2018-12-31; the target matrix contains no missing values in the current processed version. All three datasets follow a common wind-task interface with target, SCADA, NWP, and position files.

All datasets are split chronologically to avoid temporal leakage. The first 70\% of samples are used for training, the following 10\% for validation and early stopping, and the final 20\% for testing. Missing target values are filled before scaling, using forward filling followed by zero filling for residual missing entries, while original missingness can be retained by the data loader as a mask when available. Standardization parameters for power targets are fitted on the training segment and reused for validation and testing. Meteorological variables are scaled consistently across historical SCADA and forecast-horizon NWP inputs, and turbine coordinates are standardized before entering the spatial module. Calendar encodings include month, day, weekday, hour, and the native intra-hour bucket for the corresponding sampling interval.

The current repository also contains loaders and exploratory summaries for broader forecasting tasks, including GEFCom2014 Solar, GEFCom2014 Load, HK Rooftop PV, and ASHRAE building load. These datasets cover solar generation, load forecasting, rooftop PV, and building energy demand, providing additional cross-domain forecasting contexts beyond the three wind-power benchmarks emphasized in this section.

Table~\ref{tab:exp_design} summarizes the representative configurations used in the main comparison. HoT and Kelmarsh are sampled every 10 minutes, while SDWPF is sampled every 15 minutes. For 10-minute datasets, the short-term setting uses a 4-day look-back window and a 3-day forecast horizon; the medium-term setting uses about 13.3 days of history and 10 days of prediction. For SDWPF, the corresponding step counts are adjusted to its 15-minute sampling interval. This design keeps the physical duration comparable across datasets while respecting their native resolutions.

\begin{table}[!t]
\caption{Experimental Design for Representative Wind-Power Tasks.}
\label{tab:exp_design}
\centering
\scriptsize
\renewcommand{\arraystretch}{1.12}
\setlength{\tabcolsep}{3.2pt}
\begin{tabular}{lcccccc}
\toprule
\textbf{Dataset} & \textbf{$N$} & \textbf{$D_f$} & \textbf{Freq.} & \textbf{Horizon} & \textbf{$T$} & \textbf{$H$} \\
\midrule
HoT & 21 & 5 & 10 min & Short & 576 & 432 \\
HoT & 21 & 5 & 10 min & Medium & 1920 & 1440 \\
Kelmarsh & 6 & 12 & 10 min & Short & 576 & 432 \\
Kelmarsh & 6 & 12 & 10 min & Medium & 1920 & 1440 \\
SDWPF & 134 & 2 & 15 min & Short & 384 & 288 \\
SDWPF & 134 & 2 & 15 min & Medium & 1280 & 960 \\
\bottomrule
\end{tabular}
\end{table}

PhyST-Net uses a 3-layer encoder, one decoder layer, hidden size 256, feed-forward size 512, eight attention heads, dropout 0.1, and Adam optimization. Training is capped at 50 epochs with early stopping patience 5 and gradient clipping at 1.0. The loss combines mean-squared prediction error with a high-output penalty weighted by 2.0 to emphasize rated-power operating regimes. For the full model, AGSP, R-STMF, and K-AMC are enabled throughout the main comparison. The ablation scripts keep the same data split, optimizer, horizon, and model width, changing only the target module switch.

We compare PhyST-Net with strong recent time-series and spatio-temporal baselines, including TimeXer, CrossFormer, PatchTST, TiDE, iTransformer, Amplifier, DAG, GCGNet, DUET, TimeKAN, and CrossLinear. Models that do not natively consume exogenous wind variables are evaluated through the same historical-power and future-NWP protocol with a lightweight convolutional exogenous adapter, so that the comparison uses the same available information. The primary metrics are mean absolute error (MAE) and root mean square error (RMSE). Lower values indicate better predictive accuracy.

\subsection{Main Results}
\label{sec:main_results}

Table~\ref{tab:main_results} summarizes the representative benchmark results. PhyST-Net obtains the best MAE in all selected settings and the best RMSE in four out of six settings. The gains are especially pronounced on Kelmarsh, where PhyST-Net reduces MAE by 30.99\% and 25.05\% over the strongest competing baseline in the medium- and short-term tasks, respectively. On SDWPF, the advantage remains consistent, with MAE reductions of 21.44\% and 14.06\% for the medium- and short-term tasks. On HoT, PhyST-Net provides the lowest MAE in both short- and medium-term forecasting, while DAG and TimeKAN remain close competitors in RMSE. These results suggest that the proposed decomposition is most effective when the forecast must integrate non-local spatial effects and future exogenous conditions rather than relying only on near-term autoregressive continuity.

\begin{table*}[!t]
\caption{Representative Forecasting Results on Strong Settings. Best results are in bold and second-best results are underlined.}
\label{tab:main_results}
\centering
\scriptsize
\renewcommand{\arraystretch}{1.08}
\setlength{\tabcolsep}{2.0pt}
\begin{tabular}{llcccccccccccccccccccc}
\toprule
\textbf{Dataset} & \textbf{Horizon} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} & \textbf{MAE} & \textbf{RMSE} \\
 & & \multicolumn{2}{c}{\textbf{PhyST-Net}} & \multicolumn{2}{c}{\textbf{TimeXer}} & \multicolumn{2}{c}{\textbf{CrossFormer}} & \multicolumn{2}{c}{\textbf{PatchTST}} & \multicolumn{2}{c}{\textbf{iTransformer}} & \multicolumn{2}{c}{\textbf{Amplifier}} & \multicolumn{2}{c}{\textbf{DAG}} & \multicolumn{2}{c}{\textbf{DUET}} & \multicolumn{2}{c}{\textbf{TimeKAN}} \\
\cmidrule(lr){3-4}
\cmidrule(lr){5-6}
\cmidrule(lr){7-8}
\cmidrule(lr){9-10}
\cmidrule(lr){11-12}
\cmidrule(lr){13-14}
\cmidrule(lr){15-16}
\cmidrule(lr){17-18}
\cmidrule(lr){19-20}
\midrule
HoT & Medium & \textbf{267.32} & \textbf{468.23} & 377.48 & 556.27 & \underline{323.40} & 480.46 & 372.19 & 529.96 & 428.58 & 647.79 & 394.77 & 537.81 & 374.66 & 530.49 & 359.42 & 508.00 & 325.33 & \underline{477.37} \\
HoT & Short & \textbf{258.55} & \underline{449.75} & 370.98 & 536.43 & 363.96 & 526.93 & 369.48 & 504.82 & 312.79 & 496.73 & 396.91 & 549.22 & \underline{276.90} & \textbf{445.89} & 363.62 & 509.49 & 343.95 & 501.63 \\
Kelmarsh & Medium & \textbf{144.55} & \textbf{208.41} & 231.16 & 328.12 & 280.51 & 410.72 & 349.92 & 477.61 & \underline{209.45} & \underline{283.89} & 382.84 & 527.38 & 306.12 & 410.92 & 264.88 & 361.46 & 337.94 & 445.28 \\
Kelmarsh & Short & \textbf{138.67} & \textbf{200.07} & 202.08 & 285.36 & 277.38 & 397.51 & 299.90 & 404.94 & 186.61 & \underline{273.30} & 348.30 & 475.12 & \underline{185.01} & 276.33 & 290.11 & 377.49 & 329.03 & 433.20 \\
SDWPF & Medium & \textbf{129.85} & \textbf{215.72} & 198.01 & 281.86 & 185.61 & 256.27 & 249.13 & 325.73 & \underline{165.29} & \underline{233.61} & 246.56 & 350.77 & 167.78 & 246.62 & 195.13 & 277.77 & OOM & OOM \\
SDWPF & Short & \textbf{126.30} & \textbf{200.86} & 164.12 & 244.59 & 223.18 & 300.65 & 223.91 & 321.39 & 165.21 & 237.90 & 266.87 & 376.54 & \underline{146.96} & \underline{230.52} & 217.38 & 293.28 & 216.25 & 288.11 \\
\bottomrule
\end{tabular}
\vspace{2pt}
\begin{minipage}{0.98\textwidth}
\scriptsize
\textit{Notes:} OOM denotes out-of-memory failure under the same evaluation environment. The table reports representative settings in which long-range spatial interaction and future exogenous conditioning are expected to be material. Full horizon-level results are kept as supplementary evidence.
\end{minipage}
\end{table*}

\subsection{Ablation Study on SDWPF Short-Term Forecasting}
\label{sec:ablation}

We conduct the ablation study on SDWPF short-term forecasting because it provides a representative diagnostic setting with rich spatial relations and an appropriate prediction length. The 134-turbine network exposes non-trivial cross-node dependencies, while the 3-day forecast horizon is long enough for spatial propagation and future meteorological drivers to affect the forecast, but not so long that the evaluation is dominated only by accumulated uncertainty. This setting therefore offers a focused stress test for adaptive patching, graph fusion, and exogenous conditioning.

Table~\ref{tab:ablation_sdwpf_short} reports the full model and four controlled variants using the current ablation logs. The anchor model is the full PhyST-Net configuration. The no-DCP variant removes the former dynamic continuous patching branch, corresponding to the ablation of the adaptive patching component. The no-fusion variant disables the latent fusion path. The no-PSTG variant removes the former spatio-temporal graph adapter, corresponding to the R-STMF pathway. The no-K-FiLM variant removes the former exogenous modulation branch, corresponding to the K-AMC conditioning pathway. We report MAE, RMSE, $R^2$, and the grid-assessment Lambda accuracy used by the operational evaluation script.

The anchor model obtains the lowest MAE and RMSE, with a Lambda accuracy of 90.8948\%, exceeding the 83.0\% short-term assessment threshold. Removing the adaptive patching branch increases RMSE by 1.25\% and slightly reduces Lambda accuracy. Removing only the fusion path produces a smaller RMSE degradation, suggesting that the backbone preserves part of the temporal signal even without explicit fusion. Removing the spatio-temporal graph adapter increases RMSE by 0.53\%, while removing the exogenous modulation branch produces the largest RMSE degradation of 2.33\% and the lowest Lambda accuracy among the variants. These results indicate that the short-term SDWPF task benefits most from future-exogenous conditioning, while adaptive patching and relational spatial propagation provide additional measurable gains.

\begin{table}[!t]
\caption{Ablation Study on SDWPF Short-Term Forecasting.}
\label{tab:ablation_sdwpf_short}
\centering
\scriptsize
\renewcommand{\arraystretch}{1.12}
\setlength{\tabcolsep}{4.5pt}
\begin{tabular}{lccccc}
\toprule
\textbf{Variant} & \textbf{MAE} & \textbf{RMSE} & \textbf{$R^2$} & \textbf{Lambda} & \textbf{$\Delta$RMSE} \\
\midrule
PhyST-Net (anchor) & \textbf{123.139} & \textbf{201.834} & \textbf{0.7635} & \textbf{90.8948\%} & 0.00\% \\
w/o DCP & 123.851 & 204.367 & 0.7576 & 90.7713\% & +1.25\% \\
w/o fusion & 123.645 & 202.512 & 0.7619 & 90.8637\% & +0.34\% \\
w/o PSTG & 124.188 & 202.898 & 0.7610 & 90.8184\% & +0.53\% \\
w/o K-FiLM & 126.091 & 206.532 & 0.7524 & 90.5421\% & +2.33\% \\
\bottomrule
\end{tabular}
\vspace{2pt}
\begin{minipage}{0.98\linewidth}
\scriptsize
\textit{Notes:} DCP, PSTG, and K-FiLM are the historical implementation names corresponding to AGSP, R-STMF, and K-AMC, respectively. Lambda denotes the operational grid-assessment accuracy reported by the evaluation script.
\end{minipage}
\end{table}

\subsection{Discussion of the Evaluation Scope}
\label{sec:evaluation_scope}

The benchmark results show that PhyST-Net is strongest in short- and medium-horizon regimes, where future meteorological drivers and spatial propagation have sufficient time to influence the target trajectory. We therefore emphasize the operating regime aligned with the design of PhyST-Net: constrained graph forecasting under exogenous forcing.

The current ablation evidence is task-specific. It demonstrates that adaptive patching, fusion, relational graph propagation, and exogenous modulation improve SDWPF short-term forecasting, complementing the cross-dataset main comparison on SDWPF, HoT, and Kelmarsh.
