\section{Problem Definition}
\label{sec:problem_definition}
We formulate the task as ExG-aware spatio-temporal graph forecasting. Let $G=(V,E,A)$ denote a graph, where $V=\{v_1,\ldots,v_N\}$ is the node set, $E$ is the edge set, and $A\in\mathbb{R}^{N\times N}$ is an adjacency or relation matrix. A node can represent a network element, road sensor, grid bus, monitoring station, or industrial device depending on the application. The graph relation $A$ may be physical, geographic, statistical, learned, or a combination of these sources.

At forecasting origin $t$, the historical target tensor is
\begin{equation}
\mathbf{X}_{t-T+1:t}\in\mathbb{R}^{T\times N\times D_x},
\end{equation}
where $T$ is the look-back length, $N$ is the number of nodes, and $D_x$ is the target-state dimension. The future prediction target is
\begin{equation}
\mathbf{Y}_{t+1:t+H}\in\mathbb{R}^{H\times N\times D_y},
\end{equation}
where $H$ is the forecast horizon and $D_y$ is the output dimension.

We distinguish two types of exogenous variables. Historical ExG is available over the look-back window:
\begin{equation}
\mathbf{E}^{h}_{t-T+1:t}\in\mathbb{R}^{T\times D_h},
\end{equation}
and describes external or contextual conditions observed together with the target history. Future ExG is available over the forecast horizon:
\begin{equation}
\mathbf{E}^{f}_{t+1:t+H}\in\mathbb{R}^{H\times D_f},
\end{equation}
and represents known or forecastable future drivers. This separation is important because historical ExG estimates the current latent state, whereas future ExG conditions the unknown horizon. In spatio-temporal node prediction, historical ExG may include external variables and time features, while future ExG may include exogenous forecasts.

The formulation assumes an operational forecasting setting. At time $t$, the model has access to historical target states and historical ExG up to $t$, and it has access to future ExG for $t+1$ to $t+H$ only if those variables are known, scheduled, or forecast by an external source. The future target values are never used as inputs. This assumption is important for preventing leakage: variables that are measured only after the target horizon should not be included in $\mathbf{E}^{f}_{t+1:t+H}$. If an application lacks reliable future ExG, the same formulation can be used by setting $\mathbf{E}^{f}$ to calendar features, scenario forecasts, or a learned uncertainty representation, but the resulting claim should be stated as a different forecasting setting.

The graph relation is also allowed to be partially observed. In some systems, $A$ is a physical adjacency matrix; in others, it is a distance kernel, a correlation prior, or an empty prior that is refined by adaptive learning. We therefore separate $A$ from $M_G$. The matrix $A$ specifies direct relation priors used by graph propagation, while $M_G$ collects auxiliary metadata such as node coordinates, node attributes, or environmental descriptors. This separation lets the same method support both fixed-topology and adaptive-topology forecasting.


The forecasting model is a parameterized function $f_{\theta}$:
\begin{equation}
\hat{\mathbf{Y}}_{t+1:t+H}
= f_{\theta}\left(\mathbf{X}_{t-T+1:t},\mathbf{E}^{h}_{t-T+1:t},\mathbf{E}^{f}_{t+1:t+H},A,M_G\right),
\label{eq:forecast_function}
\end{equation}
where $M_G$ denotes graph metadata such as coordinates, node attributes, physical distances, or topology priors. The model should satisfy three requirements. First, it should extract temporal representations from non-stationary node histories. Second, it should propagate information across graph nodes without assuming that all spatial dependencies share one mechanism. Third, it should use future ExG only through variables available at inference time.

When domain constraints are available, the model can be decomposed into an unconstrained spatio-temporal backbone $\mathcal{H}_{\theta}$ and a constraint operator $\mathcal{C}_{\Omega}$:
\begin{equation}
\hat{\mathbf{Y}}_{t+1:t+H}
= \mathcal{C}_{\Omega}\left(\mathcal{H}_{\theta}\left(\mathbf{X}_{t-T+1:t},\mathbf{E}^{h}_{t-T+1:t},\mathbf{E}^{f}_{t+1:t+H},A,M_G\right)\right),
\label{eq:constrained_forecast}
\end{equation}
where $\Omega$ denotes domain knowledge. In spatio-temporal node prediction, $\Omega$ includes non-negativity, capacity limits, and regime feasibility. In other domains, $\Omega$ may encode speed limits, conservation constraints, bounded sensor ranges, or operational safety limits.


Given a training set $\mathcal{D}$, the supervised objective is
\begin{equation}
\min_{\theta}\; \frac{1}{|\mathcal{D}|}\sum_{(\mathbf{X},\mathbf{E}^h,\mathbf{E}^f,A,M_G,\mathbf{Y})\in\mathcal{D}}
\ell\left(\hat{\mathbf{Y}},\mathbf{Y}\right) + \lambda_{\Omega}\mathcal{R}_{\Omega}(\hat{\mathbf{Y}},\mathbf{Y}),
\label{eq:optimization_objective}
\end{equation}
where $\ell(\cdot)$ is a forecasting loss, $\mathcal{R}_{\Omega}$ is an optional constraint-aware regularizer, and $\lambda_{\Omega}$ controls its strength. The formulation does not require a specific graph operator; any module that maps temporal node states to graph-enhanced node states can be used as the graph adapter.
