\section{Related Work}
\label{sec:related_work}
This section reviews representative research trajectories pertinent to PhyST-Net. Rather than chronicling models chronologically, we categorize prior literature based on the core modeling challenges inherent to ExG-aware spatio-temporal forecasting. A summary of the architectural taxonomies is provided in Table~\ref{tab:model_comparison}.

\begin{table*}[!t]
\caption{Taxonomy and Comparative Analysis of Spatio-Temporal Forecasting Models.}
\label{tab:model_comparison}
\centering
\scriptsize
\renewcommand{\arraystretch}{1.18}
\setlength{\tabcolsep}{2.8pt}

\begin{tabularx}{\textwidth}{l l c Y Y Y Y}
\toprule
\textbf{Category} 
& \textbf{Model} 
& \textbf{Paradigm} 
& \textbf{Temporal Tokenization} 
& \textbf{Spatial Interaction} 
& \textbf{Exogenous Integration} 
& \textbf{Constraint Awareness} \\
\midrule

\multirow{4}{*}{\makecell[l]{General\\Transformer\\Forecasting}}
& Informer \cite{zhou2021informer} 
& Transformer 
& Point-wise / ProbSparse 
& -- 
& Feature concat. 
& -- \\

& Autoformer \cite{wu2021autoformer} 
& Transformer 
& Series decomposition 
& -- 
& Feature concat. 
& -- \\

& PatchTST \cite{nie2023patchtst} 
& Transformer 
& Rigid patches 
& -- 
& -- 
& -- \\

& iTransformer \cite{liu2024itransformer} 
& Transformer 
& Inverted dimension 
& Channel mixing 
& Covariate tokens 
& -- \\

\midrule

\multirow{3}{*}{\makecell[l]{Spatio-temporal\\Graph\\Forecasting}}
& STGCN \cite{yu2017spatio} 
& ST-GNN 
& Gated temporal conv. 
& Static graph conv. 
& Feature concat. 
& -- \\

& Graph WaveNet \cite{wu2019graph} 
& ST-GNN 
& Dilated causal conv. 
& Adaptive graph conv. 
& Feature concat. 
& -- \\

& SDG-Net \cite{li2024sdg} 
& ST-GNN 
& Convolutional 
& Latent dual-graph 
& -- 
& -- \\

\midrule

\multirow{5}{*}{\makecell[l]{Exogenous-aware\\Spatio-temporal\\Forecasting}}
& TimeXer \cite{wang2024timexer} 
& Hybrid 
& Rigid patches 
& Global token / cross-attn. 
& Variate tokens / cross-attn. 
& -- \\

& ConFormer \cite{conformer2025} 
& Hybrid 
& Transformer encoder 
& Cross-attention 
& External event emb. 
& -- \\

& DAG \cite{dag2025} 
& ST-GNN 
& Dependency modeling 
& Directed adaptive graph 
& ExG-guided dependency 
& -- \\

& GCGNet \cite{gcgnet2026} 
& ST-GNN 
& Generative / recurrent 
& Consistent alignment 
& Joint temporal-channel 
& -- \\

& ExoST \cite{exost2025} 
& ST-GNN 
& Convolutional 
& ExG-driven graph 
& Select-then-balance 
& -- \\

\midrule

\multirow{2}{*}{\makecell[l]{Physics-aware\\Forecasting}}
& GeatFormer \cite{liu2026geatformer} 
& Physics-inf. 
& Rigid patches 
& Dual-branch graph 
& -- 
& Penalty loss \\

& PGNN-TFT \cite{PGNNTFT2025} 
& Physics-inf. 
& Transformer / TFT 
& Static domain graph 
& -- 
& Post-hoc clipping \\

\midrule

\rowcolor{gray!12}
\textbf{Ours} 
& \textbf{PhyST-Net} 
& \textbf{Hybrid} 
& \textbf{AGSP} 
& \textbf{R-STMF} 
& \textbf{K-AMC} 
& \textbf{PI-DCH} \\

\bottomrule
\end{tabularx}

\vspace{2pt}
\begin{minipage}{\textwidth}
\scriptsize
\textit{Notes:} ``--'' indicates that the corresponding mechanism is not explicitly considered. 
ExG denotes exogenous variables; concat., emb., conv., and attn. denote concatenation, embedding, convolution, and attention, respectively. 
AGSP, R-STMF, K-AMC, and PI-DCH denote adaptive Gaussian-soft-windowed patching, relational spatio-temporal manifold fusion, KAN-based adaptive modulation conditioning, and physics-informed differentiable constraint head, respectively.
\end{minipage}

\end{table*}

\subsection{Temporal Forecasting and Transformer Architectures}
\label{sec:related_temporal}
Classical statistical forecasting techniques, including autoregressive and state-space models, retain utility when predicting low-dimensional, approximately stationary processes. However, these foundational assumptions become restrictively fragile in graph-structured systems where node dynamics are highly non-linear and persistently coupled with external forcing. While deep recurrent and convolutional architectures improved non-linear sequence modeling via latent state tracking and local temporal filtering, long-horizon forecasting remains challenging due to error accumulation and vanishing gradients over extended time steps.

Transformer-based forecasters resolve long-range dependency constraints via self-attention, establishing themselves as the dominant paradigm for deep time-series analysis. Early breakthrough models, such as Informer \cite{zhou2021informer}, Autoformer \cite{wu2021autoformer}, and FEDformer \cite{zhou2022fedformer}, focused on reducing the quadratic complexity of attention and decomposing trend-seasonal patterns. Recent variants, utilizing PatchTST-style patching and inverted-dimension attention, significantly reduce computational overhead while enhancing representation learning for multivariate sequences \cite{vaswani2017attention, nie2023patchtst, liu2024itransformer}. Nevertheless, temporal patching is predominantly implemented via rigid, fixed-length windows. In highly non-stationary sequences, physical regime transitions and rapid fluctuations rarely align with these artificial boundaries. PhyST-Net advances this trajectory by introducing AGSP, rendering temporal tokenization fully learnable and mathematically continuous.

\subsection{Spatio-Temporal Graph Forecasting}
\label{sec:related_st_graph}
Spatio-temporal graph neural networks (STGNNs) build upon foundational graph convolutions \cite{kipf2017semi} and graph attention mechanisms \cite{velickovic2018graph} to capture node interactions. These methodologies are highly effective when cross-node dependencies dictate system behavior, such as in traffic corridors and power grids \cite{yu2017spatio, wu2019graph, li2024sdg}. Furthermore, adaptive graph learning relaxes the restrictive assumption of a static, fully observable topology, allowing models to infer latent relations directly from data and dynamically adjust propagation patterns.

A critical limitation, however, is the widespread reliance on homogeneous propagation protocols across all latent states. In physical systems, spatial dependence is inherently multi-modal; nodes interact through localized proximities, shared environmental influences, and overarching exogenous drivers. A singular adjacency matrix, even if learned dynamically, conflates these distinct mechanisms. R-STMF resolves this by decomposing temporal representations into high-frequency local variations and low-frequency macro-trends, applying targeted, sparse relational propagation to each manifold prior to fusion.

\subsection{Exogenous-Variable-Aware Forecasting}
\label{sec:related_exg}
Real-world forecasting is intrinsically dependent on exogenous (ExG) variables. In vertical domains, such as intelligent traffic routing---where models like ConFormer \cite{conformer2025} leverage external factors like accidents and regulations---and resource-constrained operational forecasting, ExG features are critical for predictive resilience. 

Recent foundational architectures have begun systematically targeting ExG fusion to address inconsistent covariate effects \cite{exost2025}. In time-series modeling, TimeXer \cite{wang2024timexer} empowers Transformers by mapping external factors into variate tokens, enabling cross-attention between endogenous patches and exogenous drivers. Concurrently, DAG \cite{dag2025} introduces dual correlation networks to capture channel-wise dependencies, while in the spatio-temporal graph domain, models like GCGNet \cite{gcgnet2026} employ generative frameworks to align predicted correlations with joint temporal-channel graphs driven by external factors.

While these mechanisms significantly improve accuracy under ideal conditions, they frequently blur the temporal distinction between historical observations and future drivers. Specifically, historical ExG elucidates recent state trajectories, whereas future ExG provides horizon-level modulation. Merging both into a singular feature tensor exacerbates temporal imbalance effects \cite{exost2025} and complicates data availability reasoning at inference time. Consequently, PhyST-Net structurally decouples historical ExG from future ExG. K-AMC leverages the future-ExG pathway to generate latent modulation parameters, enabling forecast-horizon drivers to dynamically condition graph-enhanced states without being erroneously treated as historical priors.

\subsection{Interpretable Conditional Modeling with KANs}
\label{sec:related_interpretable_conditioning}
Interpretability is paramount in operational forecasting, as dispatchers require insight into the primary drivers shaping a prediction. While attention weights and post-hoc attribution methods offer diagnostic utility, they often fail to reflect the model's actual computational pathways. Kolmogorov-Arnold Networks (KAN) \cite{liu2024kan} represent a paradigm shift from traditional perceptrons by modeling transformations via learnable, univariate spline-based edge functions \cite{yamak2025review, han2024kan4tsf, vaca2024kan, xu2024kolmogorov}.

By structuring a conditioning network as an additive sum of feature-wise edge functions, the influence of individual input variables can be analytically inspected through learned basis coefficients and derivatives. PhyST-Net capitalizes on this property within the K-AMC module: future ExG features are mapped to scale and shift parameters via spline functions. While not establishing strict causal identifiability, this architectural alignment provides a transparent, inspectable explanation mechanism far superior to generic dense concatenation.

\subsection{Physics-Informed and Constraint-Aware Forecasting}
\label{sec:related_constraints}
Physics-informed learning embeds essential domain knowledge into neural architectures via structural priors, penalty losses, or projection operators. In network forecasting, enforcing such constraints is vital, as system targets are strictly bounded by operational capacities. Unfeasible predictions, regardless of aggregate error metrics, can trigger catastrophic downstream decisions.

Contemporary constraint-aware frameworks predominantly rely on physics-guided loss regularization \cite{liu2026geatformer, potosnak2024severe} or reactive post-hoc clipping \cite{welsh2025physics, PGNNTFT2025}. However, soft penalties do not guarantee compliance, and post-hoc clipping creates a structural disconnect between the trained model and inference behavior. Aligning with the philosophy of structural physics induction \cite{Shi2023UltraShort}, the PI-DCH layer internalizes feasibility operations directly within the forecasting head. By utilizing smooth barrier functions to shape predictions toward a dynamic capacity envelope prior to a final non-linear projection, the constraint pathway becomes an explicit, differentiable component of the computational graph.

\subsection{Position of This Work}
\label{sec:related_position}
PhyST-Net synthesizes advancements in Transformer forecasting, adaptive graph learning, and interpretable exogenous conditioning. Its distinguishing factor lies in the rigorous organization of the forecasting interface. Rather than presenting a bespoke, single-domain architecture, this paper formulates a generalized ExG-aware graph prediction framework, subsequently instantiating physically meaningful, domain-specific modules. This methodology positions PhyST-Net as a reusable framework for domain-constrained network forecasting.
